---
title: "Analysis of the relationship between city size and municipal spending efficiency in California cities"
subtitle: "Subtitle"
author: "Terrell Enoru"
thanks: "Project repository available at: [https://github.com/tycebot/MATH261A-project](https://github.com/tycebot/MATH261A-project)."
date: today
date-format: long
abstract: "California's rising government debt over the past decade necessitates careful examination of municipal spending efficiency. To analyze spending efficiency I use data on expenditure per capita for all california cities over the past 20 years. I use regression analysis to model the relationship between population size(the predictor) and per capita spending(the response varible). Then hypothesis testing is used on the slope to determine whether large cities benefit from economies of scale. The results were FILL THIS IN which is significant because FILL THIS IN."
format: pdf
bibliography: references.bib
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

# Data
Each observation in this dataset represents the expenditure per capita of a particular city in California at the end of a fiscal year. The key variables are entity name which is the name of the city,fiscal year(July 1 of the current year to June 30 of the next year) which is the year for which each expenditure per capita is calculated ,total expenditure which is the gross expenditure of a particular city over the fiscal year, estimated population which is the population of the city based on the Housing Unit Method which extrapolates population by taking housing units adjusting for occupation and then using that to estimate household population then combining that with estimates for "group quarters population"(essentially everyone in nontraditional households like dorms and prisons) to get the overall population estimate(more details in reference),and expenditures per capita which is the total expenditures divided by estimated population.The data is grouped by city,year, as well as binned into three groups for large medium and small sized cities.The limitations for the dataset are as follows the estimated population does a bad job of measuring homeless because of it's reliance on housing units,the years are limited to a twenty year period from 2003 to 2023,the estimated population can double count those that move until they register their new address, and the expenditures measures are both skewed over the covid period (2020-2023). None of the records are missing. By analyzing how much money is spent by a city over an extended period of time for a cities with a variety of different populations we can reasonably ascertain the relationship between expenditure and the population. However the key assumptions are that the population measurement is accurate relative to other cities in the dataset(meaning if it underestimates or overestimates the population then it does it consistently) and that the cities are correctly reporting their expenditures to the state.

per capita expenditure visualizations
```{r}

```

total expenditure visualizations
```{r}

```

estimated population visualizations
```{r}

```

# Methods
We adopt a simple linear regression model with expenditure per capita as the response and population as the predictor. Let $Y_i$ denote expenditure per capita and $Y_i$ denote population for the $Y_i$ year in our dataset. The model can be written as 
$Y_i = \beta_0 + \beta_1 X_i + \epsilon_i$
where $\beta_0$ represents the expected population when the expenditure is 0,$\beta_1$ represents the average increase in expenditure per capita for every individual added to the population, and $\epsilon_i$ represents the change in expenditure per capita not captured by the  population
This model is implemented using the lm() function from the R programming language. The lm function is the a function used to create linear models.
ADD SECTION ON HYPOTHESIS TESTING 

# Results

# References
